---
title: "Problem Set 1"
author: "Nicholas DeRobertis"
date: "03/02/26"
output: pdf_document
---

## SECTION 1: EXPLORING THE DATA

1. Load data and output first 10 rows
```{r}
brexit <- read.csv("C:/Users/nickd/EM570/Data-Storytelling-2026/assignments/Assignment1/ps1_BES.csv")
head(brexit, 10)
```

2. How many rows and columns in the dataset
```{r}
rows <- nrow(brexit)
columns <- ncol(brexit)

print(rows)
print(columns)
```

3. Print the names of the columns in the dataset 
```{r}
print(colnames(brexit))
```

4. Check what type of data is in each column
```{r}
str(brexit)
```

The current output of each column is character, character, character and integer.
Vote should be factor because each response defines a group. R cant see this without 
it being defined

Leave should be numeric, but it reads it as characters because some of it is labeled 
as missing

Education should also be numeric because of the 1-5 scale. I think it could also be factor
but I feel like numeric makes more sense for this dataset

And age should be numeric because its just respondents ages. It is listed as integer and
not numeric because each number is whole. Numeric would be output if the values we floats
or doubles

This could be fixed after cleaning the data but for now this section just explores

## SECTION 2: CLEANING THE DATA
5. How many missing or incorrectly coded values in dataset and fix them
Check if any spots are blank
```{r}
na_count <- colSums(is.na(brexit))
print(na_count)
```

Check for unique values
```{r}
lapply(brexit, unique)
```

Vote looks good
Leave and Education both have missing but other than that look fine
One instance in age showing -9

"Missing" shows up a lot, so find how many there are in each column and where
```{r}
missing_count <- sapply(brexit, function(x) sum(tolower(x) == "missing", na.rm = TRUE))
print(missing_count)
which(tolower(brexit$leave) == "missing")
```

How many times does -9 show up in age
```{r}
negative_count <- sapply(brexit, function(x) sum(x == -9, na.rm = TRUE))
print(negative_count)
which(brexit$age == -9)
```

"Missing" shows up 2852 times in leave and 3425 times in education
-9 Shows up 5 times in age
Change every "missing" and -9 to NA or blank
```{r}
clean <- function(x) {
  x[tolower(trimws(x)) == "missing"] <- NA
  x[x == -9] <- NA
  return(x)
}

brexit <- as.data.frame(sapply(brexit, clean))
```

6. If any variables are not in the appropriate format, convert them to the correct type
I mentioned issues for this before in section 1

Convert vote to factor
```{r}
brexit$vote <- as.factor(brexit$vote)
```

Convert leave to numeric
```{r}
brexit$leave <- as.numeric(brexit$leave)
```

Convert education to factor
```{r}
brexit$education <- as.numeric(brexit$education)
```

Convert age back to numeric (sapply converts everything to characters)
```{r}
brexit$age <- as.numeric(brexit$age)
str(brexit)
```

## SECTION 3: ANALYZING THE DATA
7. Calculate summary statistics for all numerical columns in the dataset
Mean for all 3 columns
```{r}
sapply(brexit[, c("leave", "education", "age")], mean, na.rm = TRUE)
```

Median for all 3 columns
```{r}
sapply(brexit[, c("leave", "education", "age")], median, na.rm = TRUE)
```

Standard Deviation for all 3 columns
```{r}
sapply(brexit[, c("leave", "education", "age")], sd, na.rm = TRUE)
```

8. Leave EU and age boxplot
```{r}
boxplot(age ~ leave, 
        data = brexit,
        main = "Leaving EU and Age Boxplot",
        xlab = "Leave Vote (0 = Stay, 1 = Leave)",
        ylab = "Age",
        col = c("lightblue", "lightgrey"))
```

9. Correlation Matrix of leave vs age
```{r}
cor_matrix <- cor(brexit[, c("age", "leave")], use = "complete.obs")
print(cor_matrix)
```

The correlation coefficient is 0.24. This means the variables are positively correlated
because the coefficient is in the positives and not negatives. Because its positively 
correlated it shows that as age increases, people are more likely to vote to leave. Even 
though this is the case, the relationship is weak because it is on the lower end of the 
max of 1.

## SECTION 4: CALCULATING DIFFERENCE IN MEANS
10. Setup
```{r}
UA <- read.csv("C:/Users/nickd/EM570/Data-Storytelling-2026/assignments/Assignment1/ps1_UA_survey.csv")
print(colnames(brexit))
```

Check for issues
```{r}
na_count_UA <- colSums(is.na(UA))
print(na_count_UA)
lapply(UA, unique)
```

Means of both tv and no tv
```{r}
mean_no_tv <- mean(UA$pro_russian_vote[UA$russian_tv == 0], na.rm = TRUE)
mean_tv <- mean(UA$pro_russian_vote[UA$russian_tv == 1], na.rm = TRUE)

print(mean_no_tv)
print(mean_tv)
```

Difference in means
```{r}
diff_means <- mean_tv - mean_no_tv
print(diff_means)
```


The mean of no tv is 17% and the mean of tv is 29%. When finding the difference in means
it proves that there's a positive relationship between respondents who receive Russian
TV broadcasts and also voting Pro-Russian.

Even though respondents who receive Russian tv broadcasts are more likely to vote Pro
Russian, this can not be considered a causal effect. This is because of a few factors.
One being that people who are already pro Russian might watch these Russian broadcasts
without it changing their views. Another thing is also in the dataset with within_25,
which shows that location plays a big role in the decision too. One way this could be
classified as a causal effect is if the tv broadcast was administered randomly to people.


